Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-03-10 text/paper2/ms.tex
add note about mixture model
more writing

\begin{abstract}
% Context
The dark-matter halo of the Milky Way is expected to be triaxial on large scales
  and filled with substructure on all scales.
It is hoped that streams or shells of stars produced by tidal disruption
  will produce very precise measures of the gravitational potential,
  to test these predictions.
% Aims
We develop a justified method for inferring the gravitational potential
  using precise measurements of stars in tidal streams
  based on an approximate generative model.
Our method does not rely on integrability of the potential;
  we do not know that any good approximation to the large-scale 
  Milky Way potential is integrable.
% Methods: 
Our model is probabilistic, with a likelihood function and priors on the parameters.
It is based on the idea that the stars in a tidal structure
  were close in phase space at some point in the past.
  Model parameters include the present-day phase-space position for each star
  and the progenitor, and thus properly accounts for finite observational uncertainties 
  and missing data.
This method advances in a few key points that have not been jointly addressed in prior work: assumptions of integrability, proper accounting for observational uncertainties, and allowing for missing dimensions.
We test our method with simulated data in a non-integrable triaxial potential, with realistic observational uncertainties for RR Lyrae stars.
Importantly, the fake data were generated with an n-body simulation,
  not with the approximate model underlying our likelihood function.
% Results:
We find that even with a small number (4 to 16) of well-measured stars,
  we can precisely infer properties of the triaxial potential,
  missing phase-space coordinates of the stars,
  and the location of the progenitor object.
The method generalizes to to encompass mixtures of independent debris structures,
  or situations in which star membership in structures is unknown.
[Get more quantitative here?]
% Conclusions: 
\end{abstract}

@@

Another approach to using halo stars as potential measures is to exploit the non-random nature of halo. For example, \citet[][LM10]{law10} used N-body simulations of the disruption of the Sgr dwarf to simultaneously model the debris and underlying Milky Way potential. By varying parameters of the Milky Way potential, they found ``best-fit'' parameters by comparing the properties of observed Sgr stars to their simulated debris over a grid of potential parameters. The computational costs of running N-body simulations limited their search to a grid of values and forced the authors to fix many other parameters (e.g., properties of the disk and bulge). Nevertheless, they were able to constrain the shape of the Milky Way potential out to $\sim$70 kpc and found that the best-fitting halo has a prolate, triaxial shape elongated in the Galactic Z direction, orthogonal to the plane of the disk. Though an unlikely orientation for the halo --- e.g., \cite{debattista13} find that the disk of the Milky Way would not remain stable in such a configuration --- LM10 showed that the data is at a state where such inference can be attempted. The computational costs associated with N-body modeling has motivated many methods that approximately model tidal streams.

@@

We identify XXX problems with the approximate models mentioned above; each method solves only a subset of these issues:
\begin{enumerate}
	\item The known debris structures are $\gtrsim$10 kpc from the Sun, thus while overdensities in the halo are less contaminated by background, the constituent stars often have incomplete or poorly measured phase-space information. Thus it is critical for any approximate model of tidal debris to incorporate observational uncertainties and missing dimensions in a consistent and justified way. 
	\item The potential of the inner galaxy is plausibly well-described by a (set of) nearly-integrable potentials, thus the dynamics of collisionless systems is conceivably captured by a transformation to some integral-of-motion-space, e.g., action-angle coordinates. However there is large uncertainty in the radial profile, shape, orientation, and graininess of the outer halo and the constancy of these parameters over distance. To test the degree to which the outer halo is well-fit by integrable potentials, it is problematic to use methods that \emph{assume} integrability. 
	\item Matching densities, small numbers of stars in Gaia %Generative models that instead rely on matching densities (in phase-space, action-space, or observed coordinates) must also include fine details about the progenitor (e.g., mass function or structure) and a model for the background of stars. 
	\item Multiple streams and other information
	\item Computational costs [...]
\end{enumerate}

%[][hereafter \paperone]
In \citet{apw13}, we introduced a simple method (\emph{Rewinder}) for using individual stars associated with tidal debris combined with knowledge about the mass and orbit of the progenitor to constrain properties of the host galaxy potential. The method exploits the relationship between the phase-space distribution of debris and (measurable) properties of the progenitor system (e.g., the tidal radius and escape velocity). In this article, we use these simple scalings --- which depend only on the mass and orbit of the progenitor and parent potential --- to present a fully probabilistic method for modeling tidal streams in any underlying potential given data with realistic uncertainties. The method relies on numerical orbit integration in ordinary phase-space and can thus incorporate arbitrarily complex models for the parent potential. The model uses individual stars as constraints --- not density matching --- and can therefore handle very small samples of stars. The model trivially extends to incorporating multiple streams, debris structures, and other kinematic information. [...something about computational cost]

In Section~\ref{sec:sims} we describe a suite of N-body simulations that span a range of progenitor masses on a characteristic, mildly eccentric orbit. In Section~\ref{sec:} we describe [...]

\section{Simulations}\label{sec:sims}
[David Hendel]
We [...] N-body simulations run with an SCF code [...]. [describe simulations here] [Describe tub]

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/potentials.pdf}
\caption{ TODO }\label{fig:potential}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/Lpts_r.png}
\caption{ Orbits of 2000 particles around $t=\tub$. The orbits were integrated backwards from their present-day positions (final time step of the N-body simulations) without the potential of the progenitor. TODO: coordinates...}\label{fig:lpts_r}
\end{center}
\end{figure}

@@

The full posterior probability for the position of a star, progenitor, and properties of the parent potential, given the data, is then
\begin{equation}
	p(\Phi, \W, \W_\sat, \tub, K, c \given \D, \D_\sat) = \frac{1}{\mathcal{Z}} p(\D \given \W) p(\D_\sat \given \W_\sat) 
												       p(\W \given \W_\sat, \tub, \Phi, K, c)
												       p(\Phi) p(\tau)p(c)
\end{equation}
where the factor $\mathcal{Z}$ only depends on the properties of the data (the \emph{evidence}), and is thus constant when varying the model parameters.

[...]

@@

\section{Experiments}
In what follows, we apply the model described above to ``observations'' of the N-body simulations described in Section~\ref{sec:sims}. For all tests in this article, we model the potential of the parent galaxy with the same functional form used in the simulations (Equation~\ref{eq:lm10}). When recovering the potential, we hold fixed the disk and spheroid parameters (see Table~\ref{tbl:potential}), along with two halo parameters: the scale radius, $r_h$, and one of the flattening components in the x-y plane, $q_2$. These halo parameters are degenerate with combinations of the other parameters --- e.g., setting $q_1=2.0$, $q_2=1.0$, $\phi=0^\circ$ is equivalent to setting $q_1=1.0$, $q_2=2.0$, $\phi=90^\circ$. The priors on the remaining halo parameters are taken to be uniform over a conservative domain of realistic values: for $v_h$, 100-200~kms$^{-1}$ corresponds to a range in solar circular velocities from $\sim$210-250 kms$^{-1}$ (holding other parameters fixed); the range in axis ratios allow for prolate, oblate, and generic triaxiality; and $\phi$ is restricted to $\pm45^\circ$ around the value measured by LM10 ($\phi = 97^\circ$).

>>>>

\begin{abstract}
% Context
The dark-matter halo of the Milky Way is expected to be triaxial on large scales
  and filled with substructure on all scales.
It is hoped that streams or shells of stars produced by tidal disruption
  will produce very precise measures of the gravitational potential,
  to test these predictions.
% Aims
We develop a justified method for inferring the gravitational potential
  using precise measurements of stars in tidal streams
  based on an approximate generative model.
Our method does not rely on integrability of the potential;
  we do not know that any good approximation to the large-scale 
  Milky Way potential is integrable.
% Methods: 
Our model is probabilistic, with a likelihood function and priors on the parameters.
It is based on the idea that the stars in a tidal structure
  were close in phase space at some point in the past.
  Model parameters include the present-day phase-space position for each star
  and the progenitor, and thus properly accounts for finite observational uncertainties 
  and missing data.
This method advances in a few key points that have not been jointly addressed in prior work: assumptions of integrability, proper accounting for observational uncertainties, and allowing for missing dimensions.
We test our method with simulated data in a non-integrable triaxial potential, with realistic observational uncertainties for RR Lyrae stars.
Importantly, the fake data were generated with an n-body simulation,
  not with the approximate model underlying our likelihood function.
% Results:
We find that even with a small number (4 to 16) of well-measured stars,
  we can precisely infer properties of the triaxial potential,
  missing phase-space coordinates of the stars,
  and the location of the progenitor object.
The method generalizes to encompass mixtures of independent debris structures,
  or situations in which star membership in structures is unknown.
% [Get more quantitative here?]
% Conclusions: 
\end{abstract}

@@

Another approach to using halo stars as potential measures is to exploit the non-random nature of halo. For example, \citet[][LM10]{law10} used n-body simulations of the disruption of the Sgr dwarf to simultaneously model the debris and underlying Milky Way potential. By varying parameters of the Milky Way potential, they found ``best-fit'' parameters by comparing the properties of observed Sgr stars to their simulated debris over a grid of potential parameters. The computational costs of running n-body simulations limited their search to a grid of values and forced the authors to fix many other parameters (e.g., properties of the disk and bulge). Nevertheless, they were able to constrain the shape of the Milky Way potential out to $\sim$70 kpc and found that the best-fitting halo has a prolate, triaxial shape elongated in the Galactic Z direction, orthogonal to the plane of the disk. Though an unlikely orientation for the halo --- e.g., \cite{debattista13} find that the disk of the Milky Way would not remain stable in such a configuration --- LM10 showed that the data is at a state where such inference can be attempted. The computational costs associated with n-body modeling has motivated many methods that approximately model tidal streams.

@@

We identify a number of problems with approximate models; each method  mentioned above solves only a subset of these issues:
\begin{enumerate}
	\item The known debris structures are $\gtrsim$10 kpc from the Sun, thus while overdensities in the halo are less contaminated by background, the constituent stars often have incomplete or poorly measured phase-space information. Thus it is critical for any approximate model of tidal debris to incorporate observational uncertainties and missing dimensions in a consistent and justified way. 
	\item The potential of the inner galaxy is plausibly well-described by a (set of) nearly-integrable potentials, thus the dynamics of collisionless systems is conceivably captured by a transformation to some integral-of-motion-space, e.g., action-angle coordinates. However there is large uncertainty in the radial profile, shape, orientation, and graininess of the outer halo and the constancy of these parameters over distance. To test the degree to which the outer halo is well-fit by integrable potentials, it is problematic to use methods that \emph{assume} integrability. 
	\item Near-future photometric surveys such as \gaia\, and LSST will likely discover many new streams and kinematic associations of stars. Potential recovery methods should be able to simultaneously model multiple streams and incorporate other dynamical constraints.
	\item Matching generated models to observed stream densities is difficult: models that rely on this must account for observational biases and properly model the underlying background of halo stars. 
	\item Full N-body models are expensive to run. Incorporating an n-body model into a general model and then performing a parameter search would be computationally intensive and, presently, intractable. 
\end{enumerate}

%[][hereafter \paperone]
In \citet{apw13}, we introduced a simple method (\emph{Rewinder}) for using individual stars associated with tidal debris combined with knowledge about the mass and orbit of the progenitor to constrain properties of the host galaxy potential. The method exploits the relationship between the phase-space distribution of debris and (measurable) properties of the progenitor system (e.g., the tidal radius and escape velocity).  Specifically, a better model for the potential is one in which orbits of test particle stars (integrated backwards from their present position) came close (within the tidal radius in position, and escape velocity in velocity) to the orbit of the progenitor. In this article, we use these simple scalings --- which depend only on the mass and orbit of the progenitor and parent potential --- to present a fully probabilistic method for modeling tidal streams in any underlying potential given data with realistic uncertainties. The method relies on numerical orbit integration in ordinary phase-space and can thus incorporate arbitrarily complex models for the parent potential. The model uses individual stars as constraints --- not density matching --- and can therefore handle very small samples of well-measured stars (e.g., 4-16). The model trivially extends to incorporating multiple streams, debris structures, and other kinematic information.

In Section~\ref{sec:sims} we describe a suite of n-body simulations that span a range of progenitor masses on a characteristic, mildly eccentric orbit. In Section~\ref{sec:} we describe [...]

\section{Simulations}\label{sec:sims}
[David Hendel]
We [...] n-body simulations run with an SCF code [...]. [describe simulations here] [Describe tub]

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/potentials.pdf}
\caption{ Equipotential contours for the LM10 potential (Eq.~\ref{eq:lm10}) in Galactocentric, cartesian coordinates for various halo parameter choices. For all panels, $v_h=121.858~\mathrm{km}/\mathrm{s}$, $r_h=12~\mathrm{kpc}$. Left to right, each column represents a new choice of parameters. If not specified, other parameters are fixed to $q_1=q_2=q_z=1$ and $\phi=0^\circ$ (far left panels).  }\label{fig:potential}
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{../../plots/paper2/Lpts_r.png}
\caption{ Orbits of 2000 particles around $t=\tub$. The orbits were integrated backwards from their present-day positions (final time step of the n-body simulations) without the potential of the progenitor. TODO: coordinates...}\label{fig:lpts_r}
\end{center}
\end{figure}

@@

This model works well when all stars are lost due to tidal pruning. Even on a mildly eccentric orbit, stars become preferentially unbound around the satellite's pericenter because of tidal shocking (cite?). We will explore properly modeling this process in future work. For this article, we modify the above to be a mixture model: we add a new parameter for each star, $P_{\rm shock}$, which is the probability that the star is drawn from a broader, ``shocked'' distribution instead of a tight Gaussian centered on an effective Lagrange point:
\begin{align}
	p_{\rm strip} &= \left [ \mathcal{N}(\bs{r} \given \bs{r}_\sat - cK\rtide \hat{x}_1, \sigma_r)\mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma_v) \left\vert\J\right\vert\right]_{t=\tub}\\
	p_{\rm shock} &= \left [ \mathcal{N}(\bs{r} \given \bs{r}_\sat, 3\sigma_r)\mathcal{N}(\bs{v} \given \bs{v}_\sat, 3\sigma_v) \left\vert\J\right\vert\right]_{t=\tub}\\
	p(\W \given \W_\sat, \tub, \Phi, K, c, P_{\rm shock}) &= P_{\rm shock}p_{\rm shock} + (1-P_{\rm shock})p_{\rm strip}
\end{align}
The full posterior probability for the position of a star, progenitor, and properties of the parent potential, given the data, is then
\begin{equation}
	p(\Phi, \W, \W_\sat, \tub, K, c \given \D, \D_\sat) = \frac{1}{\mathcal{Z}} p(\D \given \W) p(\D_\sat \given \W_\sat) 
												       p(\W \given \W_\sat, \tub, \Phi, K, c)
												       p(\Phi) p(\tau)p(c)
\end{equation}
where the factor $\mathcal{Z}$ only depends on the properties of the data (the \emph{evidence}), and is thus constant when varying the model parameters.

@@

\section{Experiments}
In what follows, we apply the model described above to ``observations'' of the n-body simulations described in Section~\ref{sec:sims}. For all tests in this article, we model the potential of the parent galaxy with the same functional form used in the simulations (Equation~\ref{eq:lm10}). When recovering the potential, we hold fixed the disk and spheroid parameters (see Table~\ref{tbl:potential}), along with two halo parameters: the scale radius, $r_h$, and one of the flattening components in the x-y plane, $q_2$. These halo parameters are degenerate with combinations of the other parameters --- e.g., setting $q_1=2.0$, $q_2=1.0$, $\phi=0^\circ$ is equivalent to setting $q_1=1.0$, $q_2=2.0$, $\phi=90^\circ$. The priors on the remaining halo parameters are taken to be uniform over a conservative domain of realistic values: for $v_h$, 100-200~km/s corresponds to a range in solar circular velocities from $\sim$210-250 km/s (holding other parameters fixed); the range in axis ratios allow for prolate, oblate, and generic triaxiality; and $\phi$ is restricted to $\pm45^\circ$ around the value measured by LM10 ($\phi = 97^\circ$).
